---
project:
  type: website
  output-dir: docs
title: "Lab 8 Machine Learning hyperparameter-tuning"
author: Jessie Magalit
format:
  html:
    self-contained: true
---

# Introduction

In this lab, we will apply the concepts we learned in Unit 3 on Machine Learning, such as feature engineering, resampling, model evaluation, and hyperparameter tuning, to a regression problem. Specifically, we will predict the `q_mean` variable using the CAMELS dataset and create a complete machine learning pipeline.

# Lab Setup

```{r setup, include=FALSE}
library(tidyverse)
library(tidymodels)
library(powerjoin)
library(skimr)
library(visdat)
library(ggpubr)
library(patchwork)
library(dials)
library(glue)

```

```{r}
root  <- 'https://gdex.ucar.edu/dataset/camels/file'

types <- c("clim", "geol", "soil", "topo", "vege", "hydro")
remote_files  <- glue('{root}/camels_{types}.txt')
local_files   <- glue('data/camels_{types}.txt')

walk2(remote_files, local_files, download.file, quiet = TRUE)

camels <- map(local_files, read_delim, show_col_types = FALSE) |> 
  power_full_join(by = 'gauge_id')
```

# Data Cleaning and EDA
```{r}
camels_clean <- camels |> 
  janitor::clean_names() |> 
  drop_na() 

skimr::skim(camels_clean)
visdat::vis_dat(camels_clean)

```

# Split the Data
```{r}
set.seed(330)

camels_split <- initial_split(camels_clean, prop = 0.8)
camels_train <- training(camels_split)
camels_test  <- testing(camels_split)

```

# Feature Engineering with Recipes
```{r}
camels_recipe <- recipe(q_mean ~ ., data = camels_train) |> 
  step_rm(gauge_lat, gauge_lon) |> 
  step_normalize(all_numeric_predictors())

```

# 1 Build Resamples
```{r}
set.seed(330)
folds <- vfold_cv(camels_train, v = 10)

```

# 2 Build 3 Candidate Models
```{r}
# Linear regression model
lm_model <- linear_reg() |> 
  set_engine("lm") |> 
  set_mode("regression")

# Decision tree model
tree_model <- decision_tree() |> 
  set_engine("rpart") |> 
  set_mode("regression")

# Random forest model
rf_model <- rand_forest() |> 
  set_engine("ranger") |> 
  set_mode("regression")


```

# 3 Test the models
```{r}
model_set <- workflow_set(
  preproc = list(camels_recipe),
  models = list(
    linear = lm_model,
    tree   = tree_model,
    forest = rf_model
  )
)

model_results <- model_set |> 
  workflow_map("fit_resamples", resamples = folds)

autoplot(model_results)


```
# 4 Model Selection
My chosen model is the random forest model because it had the lowest RMSE and MAE, and the highest R² out of the three. It seems to handle the data best and probably does well because it can pick up on complex patterns that simpler models like linear regression miss.

# Model Tuning
# 1 Build a model for your chosen specification
```{r}
# Define the random forest model with the 'trees' parameter tuned
rf_tune_model <- rand_forest(
  mtry = tune(),  # Tunable mtry parameter
  min_n = tune(), # Tunable min_n parameter
  trees = tune()  # Tunable trees parameter (added)
) |> 
  set_engine("ranger") |> 
  set_mode("regression")

```

# 2 Create a Workflow
```{r}
wf_tune <- workflow() |> 
  add_model(rf_tune_model) |> 
  add_recipe(camels_recipe)

```

# 3 Check The Tunable Values / Ranges
```{r}
# Define the hyperparameters to tune, including 'trees'
dials <- parameters(
  mtry(range = c(2, 15)),    # Number of variables to consider at each split
  min_n(range = c(2, 10)),   # Minimum number of data points in each node
  trees(range = c(50, 200))  # Number of trees to use in the forest
)

# Finalize the dials object
dials <- finalize(dials)

```

# 4 Define the Search Space
```{r}
# Assuming you are working with these hyperparameters
dials <- parameters(
  mtry(range = c(2, 15)),
  min_n(range = c(2, 10)),
  trees(range = c(50, 200))
)

# Now finalize the dials object
dials <- finalize(dials)

# Perform the grid search
set.seed(330)
my.grid <- grid_latin_hypercube(dials, size = 25)

```

# 5 Tune the Model
```{r}
# Create the Latin Hypercube grid
set.seed(330)
my.grid <- grid_latin_hypercube(dials, size = 25)

# Perform the grid search to tune the model
model_params <- tune_grid(
    wf_tune,
    resamples = folds,
    grid = my.grid,
    metrics = metric_set(rmse, rsq, mae),
    control = control_grid(save_pred = TRUE)
)

# Visualize the tuning results
autoplot(model_params)

```

# 6 Check the skill of the tuned model
```{r}
collect_metrics(model_params) |> 
  arrange(mean)

show_best(model_params, metric = "mae")

hp_best <- select_best(model_params, metric = "mae")

```

# 7 Finalize your model
```{r}
wf_final <- finalize_workflow(
  wf_tune,
  hp_best
)

```

# Final Model Verification
```{r}
final_wf <- finalize_workflow(wf_tune, hp_best)

final_fit <- last_fit(final_wf, split = camels_split)

final_predictions <- collect_predictions(final_fit)

ggplot(final_predictions, aes(x = .pred, y = q_mean)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "darkorange", linetype = "dashed") +
  geom_abline(intercept = 0, slope = 1, color = "darkgreen", linetype = "dotted") +
  labs(
    x = "Predicted q_mean",
    y = "Actual q_mean",
    title = "Predicted vs Actual q_mean",
    subtitle = "Final Random Forest Model Performance",
    caption = "Dashed = model fit | Dotted = perfect prediction"
  ) +
  theme_minimal()

```

# Building a Map!
```{r}
 # Fit finalized workflow to full cleaned dataset
final_model_fit <- fit(final_wf, camels_clean)

# Use augment() to add predictions to the original data
camels_pred <- augment(final_model_fit, new_data = camels_clean)

# Residual = (truth - prediction)^2
camels_pred <- camels_pred |>
  mutate(residual = (q_mean - .pred)^2)

# Map of predicted q_mean
map_pred <- ggplot(camels_pred, aes(x = gauge_lon, y = gauge_lat)) +
  geom_point(aes(color = .pred), size = 3) +
  scale_color_viridis_c(option = "C", name = "Predicted q_mean") +
  labs(title = "Predicted Streamflow (q_mean)") +
  theme_minimal()

# Map of residuals
map_resid <- ggplot(camels_pred, aes(x = gauge_lon, y = gauge_lat)) +
  geom_point(aes(color = residual), size = 3) +
  scale_color_viridis_c(option = "A", name = "Residual (squared)") +
  labs(title = "Prediction Residuals") +
  theme_minimal()

library(patchwork)

map_pred + map_resid

```
